What we {\sf eval} in the shadows

### 2021-04-01 paper/main.tex
sttuff

To categorize the expressions, consider a function $min(e)$ which for a given
expression $e$ returns a normal form that abstracts many incidental details
allowing the reader to focus on the structure.

We classify the resolved expressions to take into account the structure of the
expression passed to \eval. However, we do not want to discriminate expressions
based on specific values. Hence, the expression goes through several
simplification steps: constant folding of arithmetic and string expressions that
use commonly used functions (addition, string concatenation), discarding of
specific value to a symbolic value holder, variable absorption. We keep track of
variables, as they are evaluated in the environment specified to \eval, contrary
to values that are left unchanged. We also track whether assignment operators
are used in the expression or not. After the simplification steps, we obtain 15
different \emph{minimized} expressions, of which we show the eleven first most
frequent in number of sites in Table~\ref{tab:minimizedexpressions}.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
\c{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
\end{tabular}
\caption{Minimized expressions: eleven most frequent in number of sites. \#sites
  is the number of sites; \%sites the percentage of sites across the total
  number of sites (\packageNbCallSites sites) with at least this minimized
  expression; \#packages is the number of packages with at least one of the
  minimized expression and measures how widespread a minimized expression is.
  \#operations is the average number of steps performed by the R interpreter
  when given that minimized expression, and \%envir is the percentage of call
  sites for which the \c{envir} argument is explicitly passed, and is not
  \c{parent.frame()}, which is the most common explicit environment passed to
  \eval.} \label{tab:minimizedexpressions}
\end{table}

\begin{figure}[htbp]
	\centering
	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
	\caption{Most frequent combinations of minimized expressions per site,
          with a cumulative share of 95\% of the
          sites.} \label{fig:frequentcombinationminimized}
\end{figure}

\paragraph{Variables.}
The most common minimized expression is a variable.\footnote{An arithmetic
expression with some variables and constant, \eg \c{1 + x}, will also be
normalized to \c{X}, i.e. a variable.} It happens in
\packageMinimizedpropsitesa\% of the sites and is widespread across packages as
it is found in \packageMinimizedpackagea packages across our CRAN corpus.
\packageNbSymbolVarSitePercent are only symbols, \ie not complex expressions.
The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\%
of the call sites. The call sites without an explicit \c{envir} argument all use
a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
both explicit and different from \c{parent.frame()} in
\packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that when
given a variable, \eval is often used to get value in a non-local, non-parent
environment.

@@

Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$},
\texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for
instance \texttt{a\$slot + 1}. It amounts for \packageNbSlotAccess sites, \ie
\packageSlotAccessPercent of the sites. The non-default, non-parent environment
percentage is here \packageMinimizedpercentenvirnonparentg.

@@

\paragraph{Values.}
Many sites evaluates to \c{V}. \packageValOneNodePercent of those sites are only
passed an expression with one node, \ie one value, not a more complex expression
that is later constant folded. \c{V} corresponds to two main categories of
expressions passed to \eval:

\begin{itemize}
\item \eval is passed an actual value, a double, an integer, an environment, a
  closure and other simple types,
\item \eval is passed a language expression that trivially evaluates to a value,
  for instance, \c{quote(1)}, \c{quote(1 + 1)}, \quote{c("a", "b", "c")}, and
  does not include variables, so does not depend on a specific environment.
\end{itemize}

For the first category, \eval is not necessary. In
\packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie
\packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual
values. The median number of runs for those sites is
\packageMedianRunSitesUniqueActualValue (average
\packageAverageRunSitesUniqueActualValue, standard deviation
\packageSdRunSitesUniqueActualValue), suggesting it happens because of a lack of
coverage, because there are rarely triggered in our dataset and all paths are
not taken. However, manual inspection of the source code of some examples shows
that some sites actually pass values. For instance, in package \emph{ctmle}, in
function \c{evaluate\_candidate\_glmnet}:

\begin{lstlisting}
f1 <- eval(paste("A ~ ", paste(paste(names(X[,-1]), sep=''), collapse=" + ")))
f2 <- as.formula(f1)
\end{lstlisting}

\c{paste} concatenates strings and \eval will just return that string.
\c{as.formula} can take a string and convert it to a formula; \eval is not
needed here. In 20 random sites among the ones with values, a conservative
manual inspection showed that \packageUsefulValueEvalPercent of the sites had a
legitimate reason to use \eval.

The percentage of non default environment passed to \eval is
\packageMinimizedpercentenvirc when the minimized expression is a value. Some
sites which are passed a value, are also passed different minimized expressions.
For those ones, \packageNonDefaultEnvirValuePercent use a non-default
environment, and among the one that use a non-default environment,
\packageNonDefaultEnvirWithVarPercent have at least one minimized expression
with a variable inside.

@@

\paragraph{Assignments.}
This amounts to \packageAssignSitesPercent of the sites and corresponds to
minimized expressions \c{<-}, where the assignment is the top call, and
\c{F(F(<-))}, where the assignment happens inside, and others that we were not
enough present to be shown in table~\ref{tab:minimizedexpressions}. Assignments
are performed with \c{<-}, \c{<<-} or \c{assign}. Most often, assignment happens
in top call position. A non-default environment is specified only in
\packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
most of those assignments occur in the default environment, \ie in the caller of
\eval. This hints that side effects are not widespread in \eval.\footnote{Side
effects can also happen in a function call in the \eval expression, and that
function call will hide assignment operators from us. We more precisely
instrument side-effects to environment in Section~REF.}

@@

>>>>

To categorize the expressions, consider a function $min(e)$ which for a given
expression $e$ returns a normal form that abstracts many incidental details
allowing the reader to focus on the structure of the evaluated code. The
minimization function performs several simplifications steps. It performs
constant folding of arithmetic and string expressions for base operators, e.g.
$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplification. There
are 15 normal forms. We detail the interesting cases next.

\begin{itemize}
\item \underline{$min(e)=\c{V}$}: Expression that represent values are
  frequently passed to \eval. The majority of those, \packageValOneNodePercent,
  are inline constants (integers or doubles). The rest trivially evaluate to a
  value.\footnote{True as long as base functions such as \c{+} are not
  redefined. They typically aren't, but this is a limitation of this
  categorization.} In our corpus, \packageNbCallSitesUniqueActualValue call
  sites only ever see a simple values; the median number of times they are
  exercised is \packageMedianRunSitesUniqueActualValue, suggesting a possible
  lack of coverage. However, manual inspection reveals cases such as:
\begin{lstlisting}
 f1 <- eval(paste("A ~ ",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
  The code concatenates strings and \eval will simply return the string as is.
  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
  them needed to use \eval.
%% The percentage of non default environment passed to \eval is
%% \packageMinimizedpercentenvirc when the minimized expression is a value. Some
%% sites which are passed a value, are also passed different minimized expressions.
%% For those ones, \packageNonDefaultEnvirValuePercent use a non-default
%% environment, and among the one that use a non-default environment,
%% \packageNonDefaultEnvirWithVarPercent have at least one minimized expression
%% with a variable inside.

\item \underline{$min(e)=\c{X}$}: Variables are the most common normal form.
  Variables subsume values, so they include arithmetic expressions in which a
  variable occurs. \packageNbSymbolVarSitePercent of variable expression are
  symbols, \ie just an \c{x}. A variable may refer to a promise and thus trigger
  computation the first time it is accessed.
  %% The
  %% \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of
  %% the call sites. The call sites without an explicit \c{envir} argument all use
  %% a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
  %% both explicit and different from \c{parent.frame()} in
  %% \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that
  %% when given a variable, \eval is often used to get value in a non-local,
  %% non-parent environment.
\item \underline{$min(e)=\c{\$}$}: A named lookup, written \c{x\$f}, can be used
  to look up a name in an environment or a column in a data frame. This category
  also groups indexing operators, \c{[]} and \c{[[]]}. The presence of these
  operators subsumes constants and variables.
%% It amounts for \packageNbSlotAccess sites, \ie
%% \packageSlotAccessPercent of the sites. The non-default, non-parent environment
%% percentage is here \packageMinimizedpercentenvirnonparentg.

\item \underline{$min(e)=$~\c{<-}}: Assignments amount to
  \packageAssignSitesPercent of the sites, they include \c{<-}, \c{<<-} and
  \c{assign}, and subsume named lookups.
%% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
%% most of those assignments occur in the default environment, \ie in the caller of
%% \eval. This hints that side effects are not widespread in \eval.\footnote{Side
%% effects can also happen in a function call in the \eval expression, and that
%% function call will hide assignment operators from us. We more precisely
%% instrument side-effects to environment in Section~REF.}

\item \underline{$min(e)=\c{F()}$}: Function calls that

\item \underline{$min(e)=\c{F(X)}$}: Function calls that

\item \underline{$min(e)=\c{F(F(X))}$}: Function calls that

\item \underline{$min(e)=\c{FUN}$}: Function calls that

\item \underline{$min(e)=\c{\{\,\}}$}: Function calls that

\item \underline{$min(e)=\c{model.frame}$}: Function calls that

\end{itemize}

@@

@@

@@

@@

Table~\ref{tab:minimizedexpressions} gives eleven most frequent in number of
sites. \#sites is the number of sites; \%sites the percentage of sites across
the total number of sites (\packageNbCallSites sites) with at least this
minimized expression; \#packages is the number of packages with at least one of
the minimized expression and measures how widespread a minimized expression is.
\#operations is the average number of steps performed by the R interpreter when
given that minimized expression, and \%envir is the percentage of call sites for
which the \c{envir} argument is explicitly passed, and is not
\c{parent.frame()}, which is the most common explicit environment passed to
\eval.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}
